Predicting Remaining Useful Life of NASA Aircraft Engines

### 2025-02-18 articleSummaries.qmd
2/17 updates
Intro Finalization

---
title: "Literature Review - Reed"
format: html
editor: visual
self-contained: true
execute: 
  warning: false
  message: false
bibliography: references.bib
---

@@

@@

@@

>>>>

---
title: "Literature Review"
format: html
editor: visual
self-contained: true
execute: 
  warning: false
  message: false
bibliography: references.bib
---

### Article Summaries (Reed)

@@

### Article Summaries (Paton)

@@

https://link.springer.com/article/10.1186/s12891-021-04379-2

@@

https://www.researchgate.net/publication/238483310_Use_of_Schoenfeld's_global_test_to_test_the_proportional_hazards_assumption_in_the_Cox_proportional_hazards_model_An_application_to_a_clinical_study

### 2025-02-18 index.qmd
Intro Finalization

## Introduction (take 3)

@@

Cox proportional hazards (CPH) models are statistical regression models specializing in
modeling time-to-event predictions with survival data. Survival data is data with a value
for time and an event.[@Abeysekera2009] For example, in a study investigating the survival likelihood of
mechanical equipment time would begin when the equipment begins functioning and would
conclude in the event the equipment stops functioning. [@Smith2003] CPH models can use
single or multiple covariates as a means of creating predictions until failure and are
capable of dealing with survival data that contains censored data as well.[@Seung2023]
Censored data is when the information about an individual in a study is only known for a
certain period of time, such as the length of the study [@klein2005, 1]. The models also
have the advantage of being semi-parametric which allows more flexibility as compared to
other models. [@Ming-Chiang2014]

As mentioned, CPH is used in various fields including the health
industry, but the usage of the model does not come without challenges. A
COVID-19 study done using data from a Pakistani hospital noted that the
model was limited due to the recovery possibility of COVID-19 patients
as CPH assumes as time goes on, the survival probability will approach
zero with no survivors. Though CPH was not the best model for this
study, the results of the model were that individuals who were
asymptomatic and young had a higher chance of surviving the virus
[@asghar2024]. In 2024, a study was done in the US on the relationship
between chest pain occurring and mortality of the person experiencing
chest pain using CPH. A response to that study was written that provided
the limitations of CPH with the biggest concern being the assumption of
constant hazards ratio that CPH relies on for the model. In the heart
study, the effect of Triglyceride-Glucose was assumed to be a constant,
but is actually a value that fluctuates over time along with an
individual's body weight and blood glucose levels. As such, the model
would struggle to predict correctly the effect of Triglyceride-Glucose
given the violation of the constant hazards ratio [@jiang2024].

@@

## Introduction (take 2)

As the COVID-19 pandemic demonstrated, the ability to predict the
survival of a patient based on their symptoms and medical history is a
major benefit as society deals with health crisis like pandemics. Those
predictions would assist medical practitioners in triage and allow
researchers the ability to generate predictions on the individuals who
are at the greatest risk of losing their lives due to an illness. While
the medical world is the most obvious use of survival predictions, the
same methods can be applied to the predicting bank failure, the
remaining useful life of machines, machinery maintenance timelines, and
insurance likelihood of payouts. There are various survival methods that
can be examined, including Cox proportional hazards model which will be
examined in this paper.

Cox proportional hazards (CPH) model is a statistical regression model
for survival modeling that specializes in time-to-event predictions
using using single or multiple covariates and the effects they have on
survival. Traditional regression models struggle to capture the
time-to-event nature of data and treat the time variable as a fixed
value instead of continuous variable. CPH models are able to overcome
those obstacles and are capable of dealing with survival data that
contains censored data as well. Censored data is when the information
about an individual in a study is only known for a certain period of
time, such as the length of the study [@klein2005, 1]. The models also
have the advantage of being semi-parametric which allows more
flexibility as compared to other models.

As mentioned, CPH is used in various fields including the health
industry, but the usage of the model does not come without challenges. A
COVID-19 study done using data from a Pakistani hospital noted that the
model was limited due to the recovery possibility of COVID-19 patients
as CPH assumes as time goes on, the survival probability will approach
zero with no survivors. Though CPH was not the best model for this
study, the results of the model were that individuals who were
asymptomatic and young had a higher chance of surviving the virus
[@asghar2024]. In 2024, a study was done in the US on the relationship
between chest pain occurring and mortality of the person experiencing
chest pain using CPH. A response to that study was written that provided
the limitations of CPH with the biggest concern being the assumption of
constant hazards ratio that CPH relies on for the model. In the heart
study, the effect of Triglyceride-Glucose was assumed to be a constant,
but is actually a value that fluctuates over time along with an
individual's body weight and blood glucose levels. As such, the model
would struggle to predict correctly the effect of Triglyceride-Glucose
given the violation of the constant hazards ratio [@jiang2024].

An additional limitation of CPH comes from the data in the covariaties
selected and how can they can become biased. Two studies done using UK
health data demonstrated how bias can impact the values for the
covariates when relying on patient submitted information. The first was
a study on the connection between salt intake and the development of
anxiety or depression after 14.5 years [@wang2025] and the second was a
study on the association between postpartum depression and the
development of two or more chronic diseases [@zhang2025]. Both of those
studies relied on patient submitted health data and survey responses
that were supplemented with death records and other accessible health
records thus creating a possible bias in their data which could
negatively influence their model as their covariates may not have
accurately reflected the true data. These articles provide examples of
the care that needs to be taken when working with CPH models.

Continuing with using CPH outside of health data, we will study the
remaining useful life of a NASA aircraft engine using data provided by
NASA for the study on propagation modeling [@saxena2008]. From the
literature review, we have noted that we will need to be concerned with
the constant hazards ratio within our data and determine how we will
censor our data.

## Introduction (draft)

Analyzing survival data is challenging for traditional machine learning
models because they struggle to capture the time-to-event nature of the
data. Survival data typically contains challenges difficult to account
for because the datasets may be “censored” meaning that the event hasn’t
had the opportunity to occur yet by the end of the study period. Also,
traditional methods aren’t able to effectively model time-to-event data
because they treat time as fixed values opposed to treating the values
as time leading to less accurate results.

Cox proportional hazards models are commonly used to interpret, analyze,
and examine survival data. These models specialize in time-to-event
predictions and can account for multiple variables and their effects on
survival. Cox proportional hazards models are commonly used in the
health field while predicting patient survivability based on symptoms
and medical history, the financial field to predict bank failure, and
mechanical studies to predict failure of machines. Cox proportional
hazards models are popular due to their ability to incorporate multiple
covariates and they have the advantage of being semi-parametric which
allows more flexibility as compared to other models.

In this paper, we will be examining the prediction of remaining useful
life (RUL) for NASA turbofan jet engines. The lifespan of a jet engine
is important as they cost thousands of dollars to maintain and produce.
The ability to predict the remaining useful lifespan of these engines
will allow for better judgments on the finances needed to maintain or
repair them as well as alert teams that an engine may not have the RUL
for a mission they would like to use it for.

We will first generate the summary statistics for this dataset as well
as run linear and logistic regression tests. We will then test our
assumptions which include the proportional hazards assumption, the
linearity between covariates and the log hazard, that each data is
independent of each other and there is no omitted data. A model will be
generated using the survival package from R. We will then test the model
and make adjustments as needed before generating our predictions.

>>>>

## Introduction

@@

Cox proportional hazards (CPH) models are statistical regression models
specializing in modeling time-to-event predictions with survival data.
Survival data is data with a value for time and an event
[@Abeysekera2009]. For example, in a study investigating the survival
likelihood of mechanical equipment, the time value would begin when the
equipment begins functioning and would conclude in the event the
equipment stops functioning [@Smith2003]. CPH models can use single or
multiple covariates as a means of creating predictions until failure and
are capable of dealing with survival data that contains censored data as
well [@Seung2023]. Censored data is when the information about an
individual in a study is only known for a certain period of time, such
as the length of the study [@klein2005, 1]. The models also have the
advantage of being semi-parametric which allows more flexibility as
compared to other models [@Ming-Chiang2014].

As mentioned, CPH is used in various fields, primarily the health
industry, but the usage of the model does not come without challenges. A
COVID-19 study done using data from a Pakistani hospital noted that the
model was limited due to the recovery possibility of COVID-19 patients
as CPH assumes as time goes on, the survival probability will approach
zero with no survivors. Though CPH was not the best model for this
study, the results of the model were that individuals who were
asymptomatic and young had a higher chance of surviving the virus
[@asghar2024]. In 2024, a study was done in the US on the relationship
between chest pain occurring and mortality of the person experiencing
chest pain using CPH. A response to that study was written that provided
the limitations of CPH with the biggest concern being the assumption of
constant hazards ratio that CPH relies on for the model. In the heart
study, the effect of Triglyceride-Glucose was assumed to be a constant,
but is actually a value that fluctuates over time along with an
individual's body weight and blood glucose levels. As such, the model
would struggle to predict correctly the effect of Triglyceride-Glucose
given the violation of the constant hazards ratio [@jiang2024].

@@
